%~ \thispagestyle{empty}
%~ \rhead[\textsf{Volume 7, Issue 1, 2021}]{}
%~ \lhead[\textsf{INTERNATIONAL JOURNAL OF LAW SOCIAL SCIENCES}]{}

\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{footnote}{0}

\articletitle{EDITORIAL}\label{2016-editorial}


\noi
It is our immense pleasure to present International Journal of Law and Social Science, a flagship double blind peer reviewed journal of Alliance School of Law, Alliance University, Bengaluru. The IJLS 2016 having ISSN no. 2454-8553 which addresses many legal realms in order to produce conducive research and output and plays a significant role in legal research and provides a platform to Academics, Practitioners, Research Scholars and Students from all over the world contribute to the enhancement of legal concepts through this Journal. 

\noi
The main objective of this journal is to disseminate new knowledge and legal advancement to everyone in the legal arena thereby providing an open forum for the publication of high-quality original research articles on the contemporary legal issues. It examines existing and existing and emerging legal issues as well as laws as the tool for social transformation that govern them to see as to how they are evolving to fit the current situation.

\noi
The Hon’ble Chancellor and pro Vice Chancellor as well as the Editorial Bord need special thanks for their support and guidance. The Peer Review panel deserve special Applaud for their constant assistance and reviewing articles on time. We are also grateful to al those who have contributed to the publication of this journal.

\noi
To ensure journal’s future endeavor, we look forward to have more decisive contributions from academics and Research Scholars in future too. 

\begin{flushright}
Prof. Shashank Kumar\\
Editor-in-Chief\\
Assistant Professor\\
Alliance School of Law\\
Alliance University
\end{flushright}
